---
title: "Overview of seqR features"
vignette: |
  %\VignetteIndexEntry{Overview of seqR features}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}
library(seqR)
```

## Introduction

This vignette provides comprehensive overview of the features provided by `seqR` package
with appropriate examples.

The package consists of two functions that facilitate k-mer counting:

* `count_kmers` (used for counting k-mers of one type)
* `count_multimers` (a wrapper of `count_kmers`, used for counting k-mers of many types in a single invocation of the function)

and one function that facilitates processing of k-mer matrices:

* `rbind_columnwise` (a helper function used for merging several k-mer matrices that do not have same sets of columns)

## Supported input sequences

The package supports all types of biological sequences, in particular,
nucleic acids and proteins. In fact, there is no constraint regarding the domain of input.
However, input sequences must conform to the one of the following representations.

### string vectors (e.g, `c("AAA", "ACDA")`)

Example:

```{r}
count_kmers(c("AAA", "ACDA"),
            k=2)
```

This representation is more efficient and recommended to use.
However, it does not support multi-character alphabets,
as opposed to the second representation.

### list of string vectors (e.g., `list(c("A", "A", "A"), c("A", "C", "D", "A"))`)

Example:

```{r}
count_kmers(list(c("A", "A", "A"), c("A", "C", "D", "A")),
            k=2)
```

## Supported variants of k-mers

A user explicitly specifies k-mer configuration using the following function parameters:
`k`, `kmer_gaps`, and `positional` (for `count_kmers`).
There are four major variants of k-mers:
**contiguous k-mers**, **gapped k-mers**,
**positional contiguous k-mers**, and **positional gapped k-mers**.

### Contiguous k-mers

Contiguous k-mers can be defined as subwords of a fixed length.
A user specifies the length of a k-mer with `k` parameter.
The other function parameters should be `kmer_gaps = NULL` (default)
and `positional = FALSE` (default).

For example, 3-mers of sequence AABC are: AAB and ABC.

#### usage

```{r}
count_kmers(c("DDDVSVAA", "FFSFSVAAA"),
            k=5)
```

### Gapped k-mers

Gapped k-mers can be defined as subsequences (not necessarily contiguous) of a given sequence.
In particular, between two contiguous elements of the sequence there might be a gap
of the length specified by a user.

A user specifies the length of each gap separately using `kmer_gaps` parameter.
The other function parameters should be `k = length(kmer_gaps) + 1` (default)
and `positional = FALSE` (default).

For example, gapped 3-mers with gaps' lengths 0 and 1 of sequence AABCCA are:
AAC, ABC, BCA.

#### usage

```{r}
count_kmers(c("DDDVSVAA", "FFSFSVAAA"),
            kmer_gaps=c(2, 0)) # gapped 3-mer template: X__XX
```

### Positional contiguous k-mers

A positional contiguous k-mer is a subtype of a contiguous k-mer
with extra information related to the exact (start) position of the k-mer.
It means that two same contiguous k-mers that start at different positions
of a given sequence are considered to be different,
as opposed to (non-positional) contiguous k-mers.

A user specifies the length of the k-mer with `k` parameter.
The other function parameters should be `kmer_gaps = NULL` (default)
and `positional = TRUE`.

For example, positional contiguous 3-mers of sequence AABCCA are:
1_AAB, 2_ABC, 3_BCC, 4_CCA.

#### usage

```{r}
count_kmers(c("DDDVSVAA", "FFSFSVAAA"),
            k=5,
            positional=TRUE)
```

### Positional gapped k-mers

A positional gapped k-mer is a subtype of a gapped k-mer
with extra information about the exact (start) position of the k-mer.
It means that two same gapped k-mers that start at different positions
of a given sequence are considered to be different,
as opposed to (non-positional) gapped k-mers.

A user specifies the length of each gap using `kmer_gaps` parameter.
The other function parameters should be `k = length(kmer_gaps) + 1` (default)
and `positional = TRUE`.

For example, positional gapped 3-mers with gaps' lengths 0 and 1 of sequence AABCCA are:
1_AAC, 2_ABC, 3_BCA.

#### usage

```{r}
count_kmers(c("DDDVSVAA", "FFSFSVAAA"),
            kmer_gaps=c(2, 0), # gapped 3-mer template: X__XX
            positional=TRUE)
```

## Result type

The result of the k-mer counting function is a k-mer matrix,
represented as a **sparse matrix** (see [package Matrix](https://cran.r-project.org/web/packages/Matrix/index.html)).

The i-th row of the result matrix corresponds to the i-th input sequence
and contains the information about particular occurrences of k-mers.
On the other hand, columns correspond to k-mers found in input sequences
and they are represented in a human-readable form
(for more detail see section `Human-readable representation of k-mers`).

Due to the usage of a sparse matrix representation,
a signification memory optimization is achieved, as it can be seen below.

```{r, message=FALSE, warning=FALSE}
get_random_string_vector <- function(seq_num, seq_len, alphabet, seed=123) {
  set.seed(seed)
  sapply(1:seq_num,
         function(n) paste0(sample(alphabet, seq_len, replace=TRUE), collapse=""))
}

r <- count_kmers(get_random_string_vector(300, 1000, LETTERS),
                 k=4)
print(pryr::object_size(r))
print(pryr::object_size(as.matrix(r)))
```

Importantly, such a representation is supported by machine learning packages
such as [ranger](https://cran.r-project.org/web/packages/ranger/index.html)
and [xgboost](https://cran.r-project.org/web/packages/xgboost/index.html).
Therefore, in those cases, it does not require extra conversions.

## Extra features

In order to improve user experience and meet custom needs,
several extra features are provided.

### configurable k-mer alphabet

A k-mer alphabet defines which elements of a sequence
should be considered during the k-mer counting procedure.

By default, alphabet derived from input sequences is used:

```{r}
count_kmers(c("VAAAAHDFAA", "KKKKAKAAVA"),
            k=4)
```

However, a user can provide a custom k-mer alphabet (`c("A", "K")`):

```{r}
count_kmers(c("VAAAAHDFAA", "KKKKAKAAVA"),
            k=4,
            kmer_alphabet = c("A", "K"))
```

### configurable batch size

The internal algorithm processes input sequences in consecutive batches of a given size.
The larger the size of a batch is, the more sequences will be used in a single step that
processes entries concurrently.
Therefore, a user should be aware that in order to take full advantage of multi-threading,
they should set the batch size to the number that is greater or equal to the number
of CPU cores. However, simultaneously, this number should not be too large
in order not to consume too much memory, as during a single step, all sequences of a given
batch are encoded, which, in turn, increases memory usage additionally.

Generally, there are two common use cases when a user should explicitly set
the size of a batch size (in other cases, the default value is sufficient):
1. a user does not want to process sequences concurrently
(then, they should set `batch_size = 1`)
2. a user needs to tweak the memory consumption accordingly to their custom needs

#### usage

```{r}
sequences <- get_random_string_vector(seq_num = 200, seq_len = 1000, alphabet = LETTERS)
microbenchmark::microbenchmark(
  multithreaded = count_kmers(sequences, k=5, batch_size = 200),
  singlethreaded = count_kmers(sequences, k=5, batch_size = 1),
  times=11
)
```

### verbose mode

Verbose mode allows a user to get extra information about the current state of computations,
e.g., which batch of sequences is currently processed.

#### usage

```{r}
count_kmers(c("VAAAAHDFAA", "KKKKAKAAVA", "ASDDA"),
            k=3,
            batch_size=1,
            verbose = TRUE)
```

### configurable dimension of the hash value of a k-mer

As the core k-mer counting algorithm intensively takes advantage of hashing functions,
`seqR` package is a probabilistic solution.
However, due to the application of multi-dimensional hash values,
a user can specify the length of the hash vector, and, thus, increase the probability
of a correct result.

Nevertheless, for small values of `k` (`k <= 12`),
the result is certainly correct for the default value of the parameter
(i.e., `hash_size = 2`).

### compute k-mers with or without their frequencies

This feature is particularly useful when a user only wants to get information
related to presence or absence of k-mers. Then, they might set this feature
(`with_kmer_counts = FALSE`).

#### usage

Compare the following two code snippets:

```{r}
count_kmers(c("AAAAAAADFSSAAAA", "AADADADADDAD"),
            k=3,
            with_kmer_counts=FALSE)
```

```{r}
count_kmers(c("AAAAAAADFSSAAAA", "AADADADADDAD"),
            k=3,
            with_kmer_counts=TRUE)
```


### compute a result k-mer matrix with or without human-readable k-mer (column) names

The aim of this feature is to optimize memory consumption and CPU time.
It is particularly useful when many different k-mer models are tested
and there is no need to get a human-readable features.

#### usage

Compare the following two code snippets:

```{r}
count_kmers(c("AAAAAAADFSSAAAA", "AADADADADDAD"),
            k=3,
            with_kmer_names=FALSE)
```

```{r}
count_kmers(c("AAAAAAADFSSAAAA", "AADADADADDAD"),
            k=3,
            with_kmer_names=TRUE)
```

### count k-mers of several types in a single function invocation

For the sake of convenient usage,
an extra wrapper (`count_multimers`) has been implemented.
It makes possible counting k-mers of various type using only a single invocation of a function.
In this case, a user must provide sequences (rather than single values) that denote k-mer configurations
(for more detail see the documentation of `count_multimers`).

```{r}
count_multimers(c("AAAAAAADFSSAAAA", "AADADADADDAD"),
                k_vector=c(1,2,3,4),
                verbose=TRUE)
```

### merge (rbind) two k-mer matrices derived from distinct sequences

As `rbind` function does not work correctly if input matrices does not have equivalent sets of columns,
`seqR` package provides a new function (`rbind_columnwise`) to satisfy this use case:

```{r}
mA <- count_kmers(c("AAAAAAADFSSAAAA", "AADADADADDAD"),
                  k=1)
mB <- count_kmers(c("VVVVVAAVA", "ADSDDD", "AAAAAV"),
                  k=1)

rbind_columnwise(mA, mB)
```

## Human-readable representation of k-mers

Each column of the result represents a single k-mer that has the following form:

`[p_]s1.s2....sk_g1.g2...gk-1`

The first part (`[p_]`) is an integer that is used only in case of positional k-mers
and it indicates the exact (start) position of the k-mer in a sequence.
The second part (`s1.s2....sk`) represents consecutive elements of the k-mer.
Finally, the last part (`g1.g2...gk-1`) represents the consecutive lengths of gaps.
In particular, if contiguous k-mers are considered, all elements of this part is equal to 0
(e.g., `0.0.0` for 4-mers). Importantly, for 1-mers, this part is not present.
