vignette: >
  %\VignetteIndexEntry{vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Fast k-mer counting is crucial in working with peptides sequences. It may take part in building features for machine learning models for peptides prediction or as a part of sequence comparison software. SeqR provides fast k-mer counting implementation in R written with Rcpp.

## Package installation and loading

Current version is available on Github, it can be build on Windows, Linux and Mac OS. For package installation from R Console the \code{devtools} package is needed. To install seqR and load it to R session the following instructions have to be executed.

```{r setup, eval=FALSE}
#install.packages("devtools") #if you dont have it installed
devtools::install_github("piotr-ole/seqR")
library(seqR)
```

## Algorithm description

Algorithm of seqR is based on hash table that counts occurrences of each k-mer in the sequence. The vector of elements is changed to integer representation for calculation of hash function and faster code execution. K-mers can be counted within specified alphabet, elements from outside of the alphabet are not counted. Implementation also allows to count gapped k-mers. Original code was written in C++ and ported to R with Rcpp library.

## Function usage examples



## Benchmark with other software

To compare performance with other software doing k-mer counting we chose "kmer" and "biogram" packages as reference.